---
title: "DRAFT - MF algorithms for hierarchical data"
author: "Center for Veterinary Biologics - Statistics Section"
date: "November 30, 2018"
output: pdf_document
numbersections: yes
number_sections: yes
fontsize: 12pt
toc: yes
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{MF for hierarchical data} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---



```{r hiddensetup, echo = FALSE, include = FALSE}
require(kableExtra)
require(tidyverse)
require(MF)

## example data
a <- data.frame(
 room = paste('Room', rep(c('W', 'Z'), each = 24)),
 pen = paste('Pen', rep(LETTERS[1:6], each = 8)),
 litter = paste('Litter', rep(11:22, each = 4)),
 tx = rep(rep(c('vac', 'con'), each = 2), 12),
 stringsAsFactors = FALSE
 )
set.seed(76153)
a$lung[a$tx == 'vac'] <- rnorm(24, 5, 1.3)
a$lung[a$tx == 'con'] <- rnorm(24, 7, 1.3)
a <- a[-48,]
a
```

# INTRODUCTION {-}

This document is intended to provide additional details regarding the algorithms
used for calculating mitigated fraction values for hierarchical data as 
implemented in package `MF`. 

A passing familiarity with the functions `MFClusHier`, `MFnest`, 
`MFClusBootHier`, and `MFnestBoot` is helpful in understanding the examples
and underlying data assumptions in this guide. A quick start may be found at
[Quick Start for Hierarchical Data]. 

## Data assumptions. Nested Hierarchical structure. {-}

Examples in this document expect data to be structured in a *nested hierarchical 
tree*, shown in Table 1. If there are only two variables (columns) or the 
experimental design is not nested, then the methods described here may not apply.
A nested design assumes that the factor of one variable co-occur with another 
variable. For example, Pen D exists only within Room Z. If the experimental 
design of the data set is crossed (all factors of a variable co-occur with 
all factors of another variable), it is not appropriate to use this guide as 
written.

```{r, expdesign, echo = FALSE}
kable(a, align = "c", row.names = FALSE, 
  caption = "Nested hierarchical data structure.") %>%
  kable_styling('bordered', full_width = FALSE, font_size = 8) %>%  
  collapse_rows(columns = 1:4, valign = "middle")
```

\clearpage

# DETERMINING MITIGATED FRACTION {-}

## Rank tables. {-}

Deterimining the rank table is the first step of the algorithm. For each unique
level of each variable, the following values are determined: 

- `con_n` & `vac_n`: Counts of observations for each treatment (control or 
vaccinate) for a particular instance of unique factor level of a variable.
- `n1n2`: Product of counts, `con_n` * `vac_n`.
- `w`: Wilcoxon statistic.
- `u`: Mann-Whitney statistic.
- `con_medResp` & `vac_medResp`: Median observation for for each treatment level
(control or vaccinate) in a particular instance of unique factor level of a variable.

```{r, echo = FALSE}
thismfh <- MFh(formula = lung ~ tx + room/pen/litter, data = a)
thismfh
```

The structure of the rank table is the same regardless of the variable of interest.
That is, whether a user is interested in the value of mitigated fraction for variable
"pen" or "litter" does not affect the content of rows or columns, assuming the 
same input data.

### Using package MF. {-}

Rank tables can be output with the following usage in package `MF`:

- `MFClusHier(...)$MFh`
- `MFh(...)`
- `MFClusBootHier(...)$MFhBoot`
- `MFhBoot(...)`

## Summarizing the rank table. {-}

With a detailed rank table, it is possible to calculate the mitigated fraction 
for a variety of variable choices. This is the second step of the algorithm. For each variable of interest the following 
summations are calculated for each factor of that variable.

- `N1N2`: Sum of the `n1n2` values from the rank table for the factor level of 
that row.
- `U`: Sum of the `u` values from the rank table for the factor level of 
that row.
- `con_N` & `vac_N`: Sum of the counts from the rank table matching the 
particular factor level of the row. 
- `con_medResp` & `vac_medResp`: Median of responses for each comparison
group matching the particular factor level of the row. 
- `MF`: $2 * U/N1N2$.

The following example looks at the mitigated fraction of factors in variable "pen".
We can see that the values for "N1N2" have been summed from the previous rank 
table for all cases of "Pen A", "Pen B", etc.

```{r, message=FALSE, echo=FALSE}
MFnest(thismfh, which.factor = "pen")
```

It is possible to perform this summary without regard for variables to get
an overall mitigated fraction value.

```{r, message=FALSE, echo=FALSE}
MFnest(thismfh, which.factor = "All")
```

### Using package MF. {-}

- `MFClusHier(...)`: calculates rank table and MF table as a single step.
- `MFnest(...)`: calculates MF table from rank table.
- `MFClusBootHier(...)`: calculates rank table and MF table as a single step.
- `MFnestBoot(...)`: calculates MF table from rank table.


# BOOTSTRAPPING {-}

Bootstrapping can occur in one of two points of consideration: at *clusters*
or at *units*. 

## Sampling clusters. {-}
A *cluster* refers to the full grouping designation of each observation. 
For example, in Table 1 the cluster of the first observation is 
Room W/Pen A/Litter 11, and there are four observations within that cluster. 
Bootstrapping clusters samples from the unique clusters with replacement, so 
every instance will have the same number of clusters as the original data, but 
they may not be unique. The example below shows one possiblility of a bootstrap
instance from the example data. 

```{r, echo = FALSE, message = FALSE}
mfboot_multiple <- MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                                  data = a, boot.unit = FALSE, 
                                  boot.cluster = TRUE,  alpha = 0.1,
                                  which.factor = c('room', 'litter', 'All'))
mfboot_multiple$MFhBoot$bootmfh %>% 
  filter(bootID == 1) %>% 
  select(c('bootID', 'room', 'pen', 'litter'))
```

During bootstrapping of clusters, the observations within that cluster do not 
change composition. That is, Room W/Pen A/Litter 11 continues to have
two vaccinates (5.63, 4.62) and two controls (9.20, 7.28).

## Sampling units. {-}

Bootstrapping at the unit level involves sampling with replacement the observed 
values within a particular treatment group. The mix of control and vaccinate
groups remains constant. This affects the final calculation of MF because the 
`U` value may be different. 

# APPENDIX {-}

## Code used in this example {-}

```{r, eval = FALSE}
a <- data.frame(
 room = paste('Room', rep(c('W', 'Z'), each = 24)),
 pen = paste('Pen', rep(LETTERS[1:6], each = 8)),
 litter = paste('Litter', rep(11:22, each = 4)),
 tx = rep(rep(c('vac', 'con'), each = 2), 12),
 stringsAsFactors = FALSE
 )
set.seed(76153)
a$lung[a$tx == 'vac'] <- rnorm(24, 5, 1.3)
a$lung[a$tx == 'con'] <- rnorm(24, 7, 1.3)
a <- a[-48,]

# DETERMINING MITIGATED FRACTION 
thismfh <- MFh(formula = lung ~ tx + room/pen/litter, data = a)
MFnest(thismfh, which.factor = "pen")
MFnest(thismfh, which.factor = "All")

# BOOTSTRAPPING   
MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                                  data = a, boot.unit = FALSE, 
                                  boot.cluster = TRUE,  alpha = 0.1,
                                  which.factor = c('room', 'litter', 'All'))
mfboot_multiple$MFhBoot$bootmfh %>% 
  filter(bootID == 1) %>% 
  select(c('bootID', 'room', 'pen', 'litter'))
```


## Session details for this manual. {-}

```{r}
sessionInfo()
```
