---
title: "DRAFT - MF for hierarchial data"
author: "Center for Veterinary Biologics - Statistics Section"
date: "October 18, 2018"
output: pdf_document
numbersections: yes
number_sections: yes
fontsize: 12pt
toc: yes
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{MF for hierarchial data} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---


# INTRODUCTION {-}

In the specific instance of experimental data that has been organized into a hierarchial structure, that is, with two levels of treatment (e.g. control and vaccinate) and nested grouping levels (e.g. multiple litters within a single pen, multiple pens within a single room), it is not appropriate to use functions `MFClus` or `MFClusBoot` to calculate the mitigated fraction without advanced knowledge of how to filter and reshape the input data.

The new functions of `MFh`, `MFnest`, `MFClusHier`, `MFhBoot`, `MFnestBoot`, and `MFClusBootHier` allow for evaluation of the mitigated fraction dataset without the need to manipulate the input data.

**Examples**

This tutorial uses an example dataset of experimental values of the fraction of lungs with gross leisions and additional housing condition information. Example calculations include how to calculate the mitigated fraction at different variable levels, with and without bootstrapping and how to access returned values for reproducibility.

Throughout this document, we use the term *core variable* to indicate the variable corresponding to the lowest node of the hierarchial tree. *Nest variables* are those variables above the core. *All* refers to the entire tree structure.

```{r setup, message=FALSE}
require(tidyverse)
require(MF)

## example data
a <- data.frame(
 room = paste('Room',rep(c('W','Z'),each=24)),
 pen = paste('Pen',rep(LETTERS[1:6],each=8)),
 litter = paste('Litter',rep(11:22,each=4)),
 tx = rep(rep(c('vac','con'),each=2),12),
 stringsAsFactors = FALSE
 )
set.seed(76153)
a$lung[a$tx=='vac'] <- rnorm(24,5,1.3)
a$lung[a$tx=='con'] <- rnorm(24,7,1.3)
a <- a[-48,]
a
```

# Quick Start 

The two functions `MFClusHier` and `MFClusBootHier` are wrapper functions that allow calculation of MF and bootstrapped MF, respectively, as a single step.

## Calculated MF. {-}

`MFClusHier` requires four inputs:

- `formula`: The formula in form `y ~ x + a/b/c` where `y` is a continuous response, `x` is a factor with two levels of treatment, and `a/b/c` are variables corresponding to the clusters. It is expected that levels of "c" are nested within levels of "b". Nesting is assumed to be in order, left to right, highest to lowest.
- `data`: The data table (`data.frame` or `tibble`) with variables as identified in formula. If there are extra variables, they will be ignored.
- `compare`: A text vector of length two. The first position [1] is treated as the control or reference group two which members of the second position [2] are compared. If no value for the compare vector is supplied, it is assumed that the comparisons are between the `con` and `vac` treatment groups.
- `which.factor`: One or more variable(s) of interest. This can be any of the core or nest variables from the data set. If none or NULL is specified, MF will be calculated for the whole tree.

**Common Usage**

Default case, looking only at whole tree:
```{r, results = 'hide'}
mf_all <- MFClusHier(formula = lung ~ tx + room/pen/litter, data = a)
```

Selecting a specific nest variable:
```{r, results = 'hide'}
mf_room <- MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = 'room')
```

Selecting multiple variables, including the entire tree:
```{r, results = 'hide', message=FALSE}
mf_multiple <- MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = c('room', 'litter', 'All'))
```

**Output**

Output object is a list of `MFh` and `MFnest`. The first item is a rank table that can be used to reproduce MF calculations or update selection of choice for 
`which.factor`. The second item is a table that shows the MF for each unique 
level of the variable(s) specified in `which.factor`.

```{r}
mf_all
```

Within the `MFh` table the variables are:

- One row per unique core variable. The input data structure will determine 
how many variables and rows it takes to fully enumerate this.
- `con_n` & `vac_n`: Counts of observations for each treatment level in the 
particular instance of core or nest variable. Note that the left hand side of the underscore will match values passed by user to `compare`.
- `n1n2`: Product of counts, `con_n` * `vac_n`.
- `w`: Wilcoxon statistic.
- `u`: Mann-Whitney statistic.
- `con_medResp` & `vac_medResp`: Median observation for for each treatment level
in the particular instance of core or nest variable. Note that the left hand side of the underscore will match values passed by user to `compare`.

Within the `MFnest` table, the variables are:

- `variable`: Which core or nest variable is of interest for the row.
- `level`: A unique case of the variable. Input data determines how many rows it
takes to enumarate a particular variable.
- `MF`: Mitigated fraction calculated for the particular level.
- `N1N2`: Sum of the `n1n2` values from table `MFh` matching the particular 
level of interest in the row.
- `U`: Sum of the `u` values from table `MFh` matching the particular 
level of interest in the row.
- `con_N` & `vac_N`: Sum of the `_n` values from table `MFh` matching the particular level of interest in the row. Note that the left hand side of the underscore will match values passed by user to `compare`.
- `con_medResp` & `vac_medResp`: Median of responses for each comparison
group matching the particular level of interest in the row. Note that the left hand side of the underscore will match values passed by user to `compare`.

## Bootstrapped MF. {-}

`MFClusBootHier` uses all of the inputs as above plus the additional four values:

- `nboot`: Number of bootstrapping events.
- `boot.unit`: Boolean whether to sample observations from within those of the
same core.
- `boot.cluster`: Boolean whether to sample which cores are present. If TRUE,
some trees have all the cores represented in the original data while others only 
have a subset.
- `alpha`: Used to calculate empirical hpd by shortest length interval.

**Common Usage**

Default case, looking only at whole tree and bootstrapping both at the cluster
and unit levels:
```{r, results='hide'}
mfboot_all <- MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                             data = a)
```

Specifying what bootstrapping sampling to occur:
```{r, results = 'hide'}
mfboot_room <- MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                              data = a, boot.unit = TRUE, 
                              boot.cluster = FALSE, which.factor = 'room')
```

Adjusting the number of bootstrapping events and alpha:
```{r, results = 'hide', message=FALSE}
mfboot_multiple <- MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                                  data = a, boot.unit = FALSE, 
                                  boot.cluster = TRUE,  alpha = 0.1,
                                  which.factor = c('room', 'litter', 'All'))
```

**Output**

Output object is a list of `MFhBoot` and `MFnestBoot`. The former is a list
of elements that includes:

- `clusters`: The unique clusters from the original data. 
- `bootmfh`: A rank table for all the bootstrapping events. For each 
bootstrapping event, there are the number of clusters equivalent to the number of unique clusters in the original data. If boot.cluster == TRUE, a random sampling with replacment 
will be represented in each bootstrapping event. Each cluster retains 
characteristics from the original cluster, such as number of observations and 
the calculated statistics, which will be pooled when calculating the MF.
- `compare`: The values passed to compare by user.
- `mfh`: The rank table calculated from the original data.

The latter is a list of tables including:

- `mfnest_summary`: The MF for each unique level of the variable(s) specified 
in `which.factor`, including summary statistics of distribution observed by
bootstrapping.
- `mfnest_details`: The MF for each bootstrapping event for the variable(s) 
specified in `which.factor`.

```{r}
mfboot_all
```

# Reproducibility from rank tables 

It is possible to refer to the rank table to reproduce calculations of MF or to
update the choice of `which.factor`.

## Calculated MF. {-}

The function `MFnest` allows us to calculate the MF of designated nest variables 
given a rank table. If `MFClusHier` was previously called on the data, this 
information is captured in the `MFh` value:

```{r}
mf_all$MFh
```

The only other value needed is the selection passed to `which.factor`.

**Common Usage**

By default, MF for the whole tree is calculated:

```{r}
MFnest(mf_all$MFh)
```

But the user can specify one or more variables for which to evaluate MF.

```{r}
MFnest(mf_all$MFh, which.factor = 'room')
MFnest(mf_all$MFh, which.factor = c('room', 'litter', 'All'))
```

## Bootstrapped MF. {-}

The function `MFnestBoot` allows to calculate the MF of designated nest variables
given rank tables from bootstrapped and original data. If `MFClusBootHier` was 
previously called on the data, this information is captured in the `MFhBoot` value:

```{r}
mfboot_all$MFhBoot
```

The other values needed are the selection passed to `which.factor` and `alpha`.

**Common Usage**

By default, MF for the whole tree is calculated using an *alpha = 0.05*:

```{r}
MFnestBoot(mfboot_all$MFhBoot)
```

But the user can specify one or more variables for which to evaluate MF.

```{r}
MFnestBoot(mfboot_all$MFh, which.factor = 'room')
MFnestBoot(mfboot_all$MFh, which.factor = c('room', 'litter', 'All'), 
           alpha = 0.1)
```

Note that there is no sampling performed in this step. All sampling is performed
during calculation of the rank tables.

# Calculating the rank table directly 

It is also possible to calculate the rank tables independently. This might be 
useful if the rank table is needed for other calculations separate from the 
activity of calculating MF.

## Calculated {-}

The function `MFh` can be used to calculate the rank table. The usage is very 
similar to the wrapper function:

```{r}
MFh(formula = lung ~ tx + room/pen/litter, data = a)
```

To manually calculate the MF without use of wrapper function, string together 
calls to both `MFh` and `MFnest`:

```{r}
thismf <- MFh(formula = lung ~ tx + room/pen/litter, data = a)
MFnest(thismf)
```

##Bootstrapped {-}

The function `MFhBoot` can be used to calculate bootstrapped rank table. 
Usage is similar to the wrapper function:

```{r}
MFhBoot(formula = lung ~ tx + room/pen/litter, data = a, nboot = 10000,
  boot.unit = TRUE, boot.cluster = TRUE)
```

To manually calculate the MF without use of wrapper function, string together 
calls to both `MFhBoot` and `MFnestBoot`:

```{r}
thismfboot <- MFhBoot(formula = lung ~ tx + room/pen/litter, data = a)
MFnestBoot(thismfboot)
```


# Appendix
```{r}
sessionInfo()
```