---
title: "Quick Start when calculating MF from Hierarchical Data"
author: "Center for Veterinary Biologics - Statistics Section"
date: "December 21, 2018"
output: pdf_document
numbersections: yes
number_sections: yes
fontsize: 12pt
toc: yes
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Quick Start when calculating MF from Hierarchical Data} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---


```{r hiddensetup, echo = FALSE, include = FALSE}
require(kableExtra)
require(tidyverse)
require(MF)

## example data
a <- data.frame(
 room = paste('Room', rep(c('W', 'Z'), each = 24)),
 pen = paste('Pen', rep(LETTERS[1:6], each = 8)),
 litter = paste('Litter', rep(11:22, each = 4)),
 tx = rep(rep(c('vac', 'con'), each = 2), 12),
 stringsAsFactors = FALSE
 )
set.seed(76153)
a$lung[a$tx == 'vac'] <- round(rnorm(24, 5, 1.3), 2)
a$lung[a$tx == 'con'] <- round(rnorm(24, 7, 1.3), 2)
a <- a[-48,]

a
```

\clearpage

# INTRODUCTION {-}

This document is intended to supplement function help pages and 
guide users through the steps of applying functions of
the `MF` package to the evaluation of mitigated fraction when data is 
arranged in a nested hierarchical structure, first implemented in version 4.3.5. 
Details of the algorithm are covered in a separate vignette titled [Algorithms for calculating MF from hierarchical data](https://github.com/ABS-dev/MF/blob/master/inst/doc/MFHierarchicalData.pdf). When calculating MF for clustered but 
not hierarchical data, refer to help pages for functions `MFClus` and `MFClusBoot`. 
For convenience, all help pages can also be accessed via the [MF Manual](https://github.com/ABS-dev/MF/blob/master/inst/doc/MF-manual.pdf).

## Data assumptions. Nested Hierarchical structure. {-}

Examples in this document expect input data to be structured in a 
*nested hierarchical tree*, as shown in Table 1. If there is only one grouping 
variable or the experimental design is crossed, it is not appropriate to use 
this guide as written.

A nested design assumes that the factor level of one variable co-occurs with 
that of another variable. For example, Pen D exists only within Room Z. 

```{r, expdesign, echo = FALSE}
kable(a, align = "c", row.names = FALSE, 
  caption = "Nested hierarchical data structure.",
  format = 'latex') %>%
  kable_styling('bordered', full_width = FALSE, font_size = 8) %>%  
  collapse_rows(columns = 1:4, valign = "middle")
```

## Mitigated Fraction. Bootstrapped and Calculated. {-}

Users have the option of simply calculating the mitigated fraction or simultaneously
bootstrapping the mitigated fraction (including confidence interval) *and* 
calculating the mitigated fraction.
Examples of both options are shown in this document.

## Technical requirements. {-}

Examples in this manual were created using MF version `r packageVersion('MF')`, 
`r R.version.string` on Windows. CVB has not tested code usage on other systems. 

The package can be found online at: https://github.com/ABS-dev/MF/blob/master/README.md, 
including installation instructions. It is expected that users have passing familiarity 
with R code and usage, as CVB does not have resources to address training or IT
issues that may occur at external organizations.

Bug reports can be submitted online at: https://github.com/ABS-dev/MF/issues. 
To expedite resolution, please include a minimal working example and refrain 
from using confidential data. Incomplete issues may be closed without 
further investigation. Do not include confidential data as issue tracking is 
visible to all.

\clearpage

# CALCULATED MF {-}

## General use. {-}
To calculate mitigated fraction directly from data without bootstrapping, use 
the function `MFClusHier`.
This function requires four inputs:

- `formula`: The formula in form `y ~ x + a/b/c` where `y` is a continuous 
response, `x` is a factor with two levels of treatment, and `a/b/c` are grouping 
variables. Nesting is assumed to be in order, left 
to right, highest to lowest. So a single level of "a" will contain multiple 
levels of "b" and a single level of "b" will contain multiple levels of "c".
- `data`: The data table (`data.frame` or `tibble`) with variables as identified 
in formula. If there are extra variables, they will be ignored.
- `compare`: Treatment groups, a text vector of length two. The first position [1] is treated as 
the control or reference group to which members of the second position [2] are 
compared. 
- `which.factor`: One or more variable(s) of interest. This can be any of the 
grouping variables from the data set. If "All" is specified, a summary MF will be
calculated for the whole tree.

Refer to the `MFClusHier` help page (`?MFClusHier`) for extended usage details.

## Looking at MF for levels of a variable. {-}

The `which.factor` argument allows users the option of selecting if mitigated 
fraction should be calculated for levels of a particular variable. In the data
from Table 1, it may be of interest to calculate the mitigated fraction for each
level of variable "room". It is possible to consider levels from multiple 
variables simultaneously, and the designator `All` can be used to evaluate for
the entire tree, without breaking out by any variable. Evaluating the value of
mitigated fraction for the entire tree is the default behavior, should the user
not specify anything to the `which.factor` argument.

## Common coding examples. {-}

Default case, looking only at whole tree:
```{r, results = 'hide'}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a)
```

Selecting a single variable to evaluate for each factor:
```{r, results = 'hide'}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = 'room')
```

\clearpage

Selecting multiple variables, including the entire tree:
```{r, results = 'hide', message=FALSE}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = c('room', 'litter', 'All'))
```

## Basic Output. {-}

The default display of `MFClusHier` output is a table with one row
for each unique mitigated fraction calculated. A unique mitigated fraction value
will be calculated for each level of the variables identified by the user in the 
argument `which.factor` (see previous discussion). The total number of rows is the
sum of the number of unique levels for each grouping variable the user specified.
The value "All" will add one row to this table.

```{r, message = FALSE, results = 'hide'}
mf_multiple <- MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = c('room', 'litter', 'All'))
mf_multiple
```
```{r, echo=FALSE}
mf_multiple$MFnest %>% 
  mutate(MF = round(MF, 2)) %>% 
  as.data.frame
```

The table columns are:

- `variable`: Which variable was considered when evaluating MF for the row.
- `level`: A unique factor level of the variable which was considered when evaluating MF for the row.
- `MF`: Mitigated fraction calculated value.
- `N1N2`: Sum of the `n1n2` values from the rank table for the factor level of 
that row.
- `U`: Sum of the `u` values from the rank table for the factor level of 
that row.
- `con_N` & `vac_N`: Sum of the counts from the rank table matching the 
particular factor level of the row. Note that the left hand side of the underscore will match values passed by user to `compare`.
- `con_medResp` & `vac_medResp`: Median of responses for each comparison
group matching the particular factor level of the row. Note that the left hand side of the underscore will match values passed by user to argument `compare`.

## Advanced usage. {-}

This section covers the technical description of how to access the rank table
from `MFh()` and summarize using `MFnest()`. For a full discussion of the 
algorithms used in these functions and how they are related, see [Algorithms for calculating MF from hierarchical data](https://github.com/ABS-dev/MF/blob/master/inst/doc/MFHierarchicalData.pdf).

### Rank table. {-}

To access the rank table, use the **MFh** field of the output object from 
`MFClusHier()`. For example:

```{r, message = FALSE, results='hide'}
thisMFh <- mf_multiple$MFh
thisMFh
```

```{r, echo = FALSE}
thisMFh$coreTbl %>%
  as.data.frame
```

Alternatively, calculate rank table directly by using the `MFh()` function:

```{r, message=FALSE, eval=FALSE}
MFh(formula = lung ~ tx + room/pen/litter, data = a)
```
```{r, message=FALSE, echo=FALSE}
MFh(formula = lung ~ tx + room/pen/litter, data = a)$coreTbl %>% 
  as.data.frame(.)
```

The output table includes the following information, used to calculate the
Basic Output table as described previously:

- `con_n` & `vac_n`: Counts of observations for each treatment for a 
particular instance of unique factor level of a variable. Note that the left 
hand side of the underscore will match values passed by user to `compare`.
- `n1n2`: Product of counts, `con_n` * `vac_n`.
- `w`: Wilcoxon statistic.
- `u`: Mann-Whitney statistic.
- `con_medResp` & `vac_medResp`: Median observed response for each treatment group
in a particular instance of unique factor level. Note that the 
left hand side of the underscore will match values passed by user to `compare`.

There is one row for each unique combination of factor levels across
all variables. This table shows the initial statistics as determined by 
the experimental design. Refer to the vignette for algorithm design
[Algorithms for calculating MF from hierarchical data](https://github.com/ABS-dev/MF/blob/master/inst/doc/MFHierarchicalData.pdf) 
for in-depth discussion of how the 
rank is used to evaluate mitigated fraction values.

The rank table is not affected by changes to the `which.factor` argument.

### Reproducibility. {-}

The rank table can be used to re-calculate the mitigated fraction values, for 
example if the user intends to explore a different selection to the 
`which.factor` argument. To do this, use the function `MFnest` which takes
the following input arguments:

- `Y`: **MFh** field output as from `MFClusHier()` 
- `which.factor`: As above.

For example:

```{r, message=FALSE, eval=FALSE}
MFnest(thisMFh, which.factor = "pen")
```

```{r, message=FALSE, echo=FALSE}
MFnest(thisMFh, which.factor = "pen") %>% as.data.frame
```

This is the same output as if the user had initially selected for variable "pen":

```{r, message=FALSE, eval=FALSE}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = "pen")
```
```{r, message=FALSE, echo=FALSE}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = "pen")$MFnest %>% 
  as.data.frame
```
# BOOTSTRAPPED MF {-}

## General use. {-}

The function `MFClusBootHier` allows for a bootstrapping approach to calculating
mitigated fraction values. Input arguments are:

*Same as in non-bootstrapped usage (i.e. MFClusHier)*

- `formula`: The formula in form `y ~ x + a/b/c` where `y` is a continuous 
response, `x` is a factor with two levels of treatment, and `a/b/c` are grouping 
variables. Nesting is assumed to be in order, left 
to right, highest to lowest. So a single level of "a" will contain multiple 
levels of "b" and a single level of "b" will contain multiple levels of "c".
- `data`: The data table (`data.frame` or `tibble`) with variables as identified in argument `formula`. If there are extra variables, they will be ignored.
- `compare`: Treatment groups, a text vector of length two. The first position [1] is treated as the control or reference group to which members of the second position [2] are compared. 
- `which.factor`: Variable(s) of interest. This can be any of the 
grouping variables from the data set. If "All" is specified, a summary MF will be
calculated for the whole tree.

*Additional arguments*

- `nboot`: Number of bootstrapping events.
- `boot.unit`: Boolean whether to sample observations from within those of the
same core.
- `boot.cluster`: Boolean whether to sample which clusters are present. If TRUE,
some trees have all the clusters represented in the original data while others only 
have a subset.
- `alpha`: Complement of the confidence level. As used in `MFClusBoot`.
- `seed`: Used to initialize random number generator for reproducibility.

A "core" is the unique combination of variable levels from the data, including
the `compare` designation. In Table 1, one core would 
be Room W/Pen A/Litter 11/vac and another would be Room Z/Pen F/Litter 22/con.

A "cluster" is the the unique combination of variable levels from the data, 
without the `compare` designation.  In Table 1, the first four observations 
are from the same core, Room W/Pen A/Litter 11.

Refer to the `MFClusHier` help page (`?MFClusBootHier`) for extended usage details.

Further discussion regarding the bootstrapping algorithm can be found in 
[Algorithms for calculating MF from hierarchical data](https://github.com/ABS-dev/MF/blob/master/inst/doc/MFHierarchicalData.pdf). 

## Common coding examples. {-}

Default case, looking only at whole tree and bootstrapping both at the cluster
and unit levels:

```{r, results='hide'}
MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                             data = a)
```

Specifying what bootstrapping sampling to occur:
```{r, results = 'hide'}
MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                              data = a, boot.unit = TRUE, 
                              boot.cluster = FALSE, which.factor = 'room')
```

Adjusting the number of bootstrapping events and alpha:
```{r, results = 'hide', message=FALSE}
MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                                  data = a, boot.unit = FALSE, 
                                  boot.cluster = TRUE,  alpha = 0.1,
                                  which.factor = c('room', 'litter', 'All'))
```

\clearpage

## Basic output. {-}

The default display of `MFClusBootHier` is a table with one row for each unique
mitigated fraction calculated, just like in the non-bootstrapped approach. 
However, instead of summary statistics about the calculated mitigated fraction,
there are values summarizing the mitigated fraction from a bootstrapped
population.

```{r, message = FALSE, results='hide'}
mfboot_multiple <- MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                                  data = a, boot.unit = FALSE, 
                                  boot.cluster = TRUE,  alpha = 0.1,
                                  which.factor = c('room', 'litter', 'All'),
                                  seed = 150)
mfboot_multiple
```

```{r, message=FALSE, echo=FALSE}
mfboot_multiple$MFnestBoot$mfnest_summary %>%
  mutate(median = round(median, 2),
    etlower = round(etlower, 2),
    mf.obs = round(mf.obs, 2)) %>%
  as.data.frame
```

\clearpage

The variable columns are:

- `variable`: Which variable was considered when evaluating MF for the row.
- `level`: A unique factor level of the variable which was considered when 
evaluating MF for the row.
- `median`: Median of mitigated fractions calculated from the bootstrapped 
population.
- `etlower`: Lower value of equal tailed range of mitigated fractions calculated 
from the bootstrapped population.
- `etupper`: Upper value of equal tailed range of mitigated fractions calculated 
from the bootstrapped population.
- `hdlower`: Lower value of the highest posterior density range of mitigated fractions calculated 
from the bootstrapped population.
- `hdupper`: Upper value of the highest posterior density range of mitigated fractions calculated 
from the bootstrapped population.
- `mf.obs`: Mitigated fraction value calculated from data input, without 
bootstrapping.

## Advanced usage. {-}

This section covers the technical description of how to access the bootstrapped 
results and use the `MFnestBoot()` function. For a full discussion of the 
algorithms of how `MFhBoot()`, 
`MFnestBoot()` and `MFClusBootHier` are related, or details of the bootstrapping
algorithm, see [Algorithms for calculating MF from hierarchical data](https://github.com/ABS-dev/MF/blob/master/inst/doc/MF-manual.pdf).

### Bootstrapping step output. {-}

The bootstrapping stage is the most computationally intensive, so a user may
wish to bypass this step subsequently if the only change are values 
being passed to the `which.factor` or `alpha` arguments. Access bootstrapping step outputs
using the **MFhBoot** field of the output object. For example:

```{r, message=FALSE}
thisBootMFh <- mfboot_multiple$MFhBoot
thisBootMFh
```

It is possible to calculate `mfboot_multiple$MFhBoot` directly by using the 
`MFhBoot()` function, however this executes the bootstrapping stage again. To
get the same output with both approaches, the `seed` argument must be the same
value.

```{r, message=FALSE, results=FALSE}
MFhBoot(formula = lung ~ tx + room/pen/litter, 
                                    data = a, boot.unit = FALSE, 
                                    boot.cluster = TRUE,  
                                    seed = 150)
```

This output is a list of four objects:

- `bootmfh`: The rank table of all the bootstrapped data sets. This is 
formatted as the rank table for non-bootstrapped data with designation of a single
bootstrapping instance using the additional variable **bootID**. Values 
for **bootID** are simply `1:nboot`.
- `clusters`: Unique clusters in the data.
- `compare`: User-supplied value passed to compare argument.
- `mfh`: The rank table from user-supplied data; formatted as for non-bootstrapped data.

Note that in the particular case of `boot.cluster = FALSE`, although all 
`nboot` instances will be the same clusters as input data, observed responses 
will vary due to sampling of the cores. The case of both
`boot.cluster` and `boot.unit` as FALSE is simply calculating MF from data 
without any bootstrapping.

### Reproducibility. {-}

To bootstrap from the same data in a reproducible manner, numerical 
value for the **seed** argument in `MFClusBootHier`. If no value for `seed` is 
specified, the function defaults to a value of `sample(1:1e+05, 1)` and 
subsequent iterations may yield different summaries.

```{r, results='hide'}
MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                              data = a, seed = 150)
```

Since the bootstrapping step is the most computationally intensive, a user may
wish to use the bootstrapped rank table to recalculate mitigated fraction for 
levels  of different variables or a different `alpha` for calculating confidence 
intervals. This is possible using the `MFnestBoot` function,
which takes the following input arguments:

- `x`: **MFhBoot** field output as from `MFClusBootHier`.
- `which.factor`: As above.
- `alpha`: As above.

For example:

```{r}
MFnestBoot(thisBootMFh, which.factor = c('pen', 'All'), alpha = 0.1)
```

# APPENDIX {-}

## Code for example data. {-}

```{r, eval=FALSE}
a <- data.frame(
 room = paste('Room', rep(c('W', 'Z'), each = 24)),
 pen = paste('Pen', rep(LETTERS[1:6], each = 8)),
 litter = paste('Litter', rep(11:22, each = 4)),
 tx = rep(rep(c('vac', 'con'), each = 2), 12),
 stringsAsFactors = FALSE
 )
set.seed(76153)
a$lung[a$tx == 'vac'] <- round(rnorm(24, 5, 1.3), 2)
a$lung[a$tx == 'con'] <- round(rnorm(24, 7, 1.3), 2)
a <- a[-48,]
```

## Session details for this manual. {-}

```{r}
sessionInfo()
```




