---
title: "DRAFT - Quick Start for Hierarchical Data"
author: "Center for Veterinary Biologics - Statistics Section"
date: "November 30, 2018"
output: pdf_document
numbersections: yes
number_sections: yes
fontsize: 12pt
toc: yes
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Quick Start for Hierarchical Data} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---


```{r hiddensetup, echo = FALSE, include = FALSE}
require(kableExtra)
require(tidyverse)
require(MF)

## example data
a <- data.frame(
 room = paste('Room', rep(c('W', 'Z'), each = 24)),
 pen = paste('Pen', rep(LETTERS[1:6], each = 8)),
 litter = paste('Litter', rep(11:22, each = 4)),
 tx = rep(rep(c('vac', 'con'), each = 2), 12),
 stringsAsFactors = FALSE
 )
set.seed(76153)
a$lung[a$tx == 'vac'] <- rnorm(24, 5, 1.3)
a$lung[a$tx == 'con'] <- rnorm(24, 7, 1.3)
a <- a[-48,]
a
```

# INTRODUCTION {-}

This document is intended to guide users through initial usage of functions in 
the `MF` package for evaluation of mitigated fraction when data is arranged in 
a nested hierarchical structure. Details of the algorithm are covered in a 
separate vignette.

## Data assumptions. Nested Hierarchical structure. {-}

Examples in this document expect data to be structured in a *nested hierarchical 
tree*, shown in Table 1. If there are only two variables (columns) or the 
experimental design is not nested, then the methods described here may not apply.
A nested design assumes that the factor of one variable co-occur with another 
variable. For example, Pen D exists only within Room Z. If the experimental 
design of the data set is crossed (all factors of a variable co-occur with 
all factors of another variable), it is not appropriate to use this guide as 
written.

```{r, expdesign, echo = FALSE}
kable(a, align = "c", row.names = FALSE, 
  caption = "Nested hierarchical data structure.") %>%
  kable_styling('bordered', full_width = FALSE, font_size = 8) %>%  
  collapse_rows(columns = 1:4, valign = "middle")
```

## Mitigated Fraction. Bootstrapped and Calculated. {-}

Users have the option of directly calculating the mitigated fraction or 
bootstrapping the mitigated fraction value. Examples of both options are shown
in this document.

## MF package basics. {-}

Examples in this document make use of CVB-authored package `MF`, which can
be found online at: https://github.com/ABS-dev/MF/blob/master/README.md, including
installation instructions. It is expected that users have passing familiarty 
with R code and usage, as CVB does not have resources to address training or IT
issues that may occur at external organizations.

Bug reports can be submitted online at: https://github.com/ABS-dev/MF/issues. 
To expedite resolution, please include a minimal working example and refrain 
from using confidential data. Incomplete issues may be closed without 
further investigation. Questions specific to a submission should be directed
to the appropriate Reviewer.

# CALCULATED MF {-}

To calculate mitigated fraction directly from data, use the function `MFClusHier`.
This function requires four inputs:

- `formula`: The formula in form `y ~ x + a/b/c` where `y` is a continuous response, `x` is a factor with two levels of treatment, and `a/b/c` are variables corresponding to the clusters. It is expected that levels of "c" are nested within levels of "b". Nesting is assumed to be in order, left to right, highest to lowest.
- `data`: The data table (`data.frame` or `tibble`) with variables as identified in formula. If there are extra variables, they will be ignored.
- `compare`: A text vector of length two. The first position [1] is treated as the control or reference group two which members of the second position [2] are compared. If no value for the compare vector is supplied, it is assumed that the comparisons are between the `con` and `vac` treatment groups.
- `which.factor`: One or more variable(s) of interest. This can be any of the variables from the data set. If none or NULL is specified, MF will be calculated for the whole tree.

## Looking at MF for factors of a variable. {-}

The `which.factor` argument allows users the option of selecting if mitigated 
fraction should be calculated for factors of a particular variable. In the data
from Table 1, it may be of interest to calculate the mitigated fraction for each
factor of variable "room". It is possible to consider factors from multiple 
variables simultaneously, and the designator `All` can be used to evaluate for
the entire tree, without breaking out by any variable. Evaluating the value of
mitigated fraction for the entire tree is the default behavior, should the user
not specify anything to the `which.factor` argument.

## Common coding examples. {-}

Default case, looking only at whole tree:
```{r, results = 'hide'}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a)
```

Selecting a single variable to evaluate for each factor:
```{r, results = 'hide'}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = 'room')
```

Selecting multiple variables, including the entire tree:
```{r, results = 'hide', message=FALSE}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = c('room', 'litter', 'All'))
```

## Output. {-}

When the function `MFClusHier` is first run, it will display a table with one row
for each unique mitigated fraction calculated. A unique mitigated fraction value
will be calculated for each factor of the variables identified by the user in the 
argument `which.factor` (see previous discussion).

```{r, message = FALSE}
mf_multiple <- MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = c('room', 'litter', 'All'))
```