---
title: "DRAFT - Quick Start for Hierarchical Data"
author: "Center for Veterinary Biologics - Statistics Section"
date: "November 30, 2018"
output: pdf_document
numbersections: yes
number_sections: yes
fontsize: 12pt
toc: yes
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{Quick Start for Hierarchical Data} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---


```{r hiddensetup, echo = FALSE, include = FALSE}
require(kableExtra)
require(tidyverse)
require(MF)

## example data
a <- data.frame(
 room = paste('Room', rep(c('W', 'Z'), each = 24)),
 pen = paste('Pen', rep(LETTERS[1:6], each = 8)),
 litter = paste('Litter', rep(11:22, each = 4)),
 tx = rep(rep(c('vac', 'con'), each = 2), 12),
 stringsAsFactors = FALSE
 )
set.seed(76153)
a$lung[a$tx == 'vac'] <- rnorm(24, 5, 1.3)
a$lung[a$tx == 'con'] <- rnorm(24, 7, 1.3)
a <- a[-48,]
a
```

\clearpage

# INTRODUCTION {-}

This document is intended to guide users through initial usage of functions in 
the `MF` package for evaluation of mitigated fraction when data is arranged in 
a nested hierarchical structure. Details of the algorithm are covered in a 
separate vignette.

## Data assumptions. Nested Hierarchical structure. {-}

Examples in this document expect data to be structured in a *nested hierarchical 
tree*, shown in Table 1. If there are only two variables (columns) or the 
experimental design is not nested, then the methods described here may not apply.
A nested design assumes that the factor of one variable co-occur with another 
variable. For example, Pen D exists only within Room Z. If the experimental 
design of the data set is crossed (all factors of a variable co-occur with 
all factors of another variable), it is not appropriate to use this guide as 
written.

```{r, expdesign, echo = FALSE}
kable(a, align = "c", row.names = FALSE, 
  caption = "Nested hierarchical data structure.",
  format = 'latex') %>%
  kable_styling('bordered', full_width = FALSE, font_size = 8) %>%  
  collapse_rows(columns = 1:4, valign = "middle")
```

## Mitigated Fraction. Bootstrapped and Calculated. {-}

Users have the option of directly calculating the mitigated fraction or 
bootstrapping the mitigated fraction value. Examples of both options are shown
in this document.

## MF package basics. {-}

Examples in this document make use of CVB-authored package `MF`, which can
be found online at: https://github.com/ABS-dev/MF/blob/master/README.md, including
installation instructions. It is expected that users have passing familiarity 
with R code and usage, as CVB does not have resources to address training or IT
issues that may occur at external organizations.

Bug reports can be submitted online at: https://github.com/ABS-dev/MF/issues. 
To expedite resolution, please include a minimal working example and refrain 
from using confidential data. Incomplete issues may be closed without 
further investigation. Questions specific to a submission should be directed
to the appropriate Reviewer.

## Technical requirements. {-}

Examples in this manual were created using MF version `r packageVersion('MF')`, 
`r R.version.string` on Windows. CVB has not tested code usage on other systems. 

\clearpage

# CALCULATED MF {-}

## General use. {-}
To calculate mitigated fraction directly from data, use the function `MFClusHier`.
This function requires four inputs:

- `formula`: The formula in form `y ~ x + a/b/c` where `y` is a continuous response, `x` is a factor with two levels of treatment, and `a/b/c` are variables corresponding to the clusters. It is expected that levels of "c" are nested within levels of "b". Nesting is assumed to be in order, left to right, highest to lowest.
- `data`: The data table (`data.frame` or `tibble`) with variables as identified in formula. If there are extra variables, they will be ignored.
- `compare`: A text vector of length two. The first position [1] is treated as the control or reference group two which members of the second position [2] are compared. If no value for the compare vector is supplied, it is assumed that the comparisons are between the `con` and `vac` treatment groups.
- `which.factor`: One or more variable(s) of interest. This can be any of the variables from the data set. If none or NULL is specified, MF will be calculated for the whole tree.

## Looking at MF for factors of a variable. {-}

The `which.factor` argument allows users the option of selecting if mitigated 
fraction should be calculated for factors of a particular variable. In the data
from Table 1, it may be of interest to calculate the mitigated fraction for each
factor of variable "room". It is possible to consider factors from multiple 
variables simultaneously, and the designator `All` can be used to evaluate for
the entire tree, without breaking out by any variable. Evaluating the value of
mitigated fraction for the entire tree is the default behavior, should the user
not specify anything to the `which.factor` argument.

## Common coding examples. {-}

Default case, looking only at whole tree:
```{r, results = 'hide'}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a)
```

Selecting a single variable to evaluate for each factor:
```{r, results = 'hide'}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = 'room')
```

\clearpage

Selecting multiple variables, including the entire tree:
```{r, results = 'hide', message=FALSE}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = c('room', 'litter', 'All'))
```

## Basic Output. {-}

The default display of `MFClusHier` output is a table with one row
for each unique mitigated fraction calculated. A unique mitigated fraction value
will be calculated for each factor of the variables identified by the user in the 
argument `which.factor` (see previous discussion). The total number of rows is 
sum of the number of unique factors for each variable the user specified.

```{r, message = FALSE}
mf_multiple <- MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = c('room', 'litter', 'All'))
mf_multiple
```

The variable columns are:

- `variable`: Which variable was considered when evaluating MF for the row.
- `level`: A unique factor level of the variable which was considered when evaluating MF for the row.
- `MF`: Mitigated fraction calculated value.
- `N1N2`: Sum of the `n1n2` values from the rank table for the factor level of 
that row.
- `U`: Sum of the `u` values from the rank table for the factor level of 
that row.
- `con_N` & `vac_N`: Sum of the counts from the rank table matching the 
particular factor level of the row. Note that the left hand side of the underscore will match values passed by user to `compare`.
- `con_medResp` & `vac_medResp`: Median of responses for each comparison
group matching the particular factor level of the row. Note that the left hand side of the underscore will match values passed by user to `compare`.

## Advanced usage. {-}

This section covers the technical description of how to access **MFh** and `MFnest()`
function. For a full discussion of the algorithms of how **MFh**, `MFnest()` and 
`MFClusHier` are related, see [MF algorithms for hierarchical data].

### Rank table. {-}

To access the rank table, use the **MFh** field of the output object. For example:

```{r, message = FALSE}
thisMFh <- mf_multiple$MFh
thisMFh
```

This table includes the following information, used in the calculation of the
basic output table described previously:

- `con_n` & `vac_n`: Counts of observations for each treatment for a 
particular instance of unique factor level of a variable. Note that the left 
hand side of the underscore will match values passed by user to `compare`.
- `n1n2`: Product of counts, `con_n` * `vac_n`.
- `w`: Wilcoxon statistic.
- `u`: Mann-Whitney statistic.
- `con_medResp` & `vac_medResp`: Median observation for for each treatment level
in a particular instance of unique factor level of a variable. Note that the 
left hand side of the underscore will match values passed by user to `compare`.

There is one row for each unique combination of variable factor levels across
the whole data set. This table shows the initial statistics as determined by 
the experimental design. Refer to the vignette for algorithm design
[MF algorithms for hierarchical data] for in-depth discussion of how the 
rank is used to evaluate mitigated fraction values.

### Reproducibility. {-}

The rank table can be used to re-calculate the mitigated fraction values, for 
example if the user intends to explore a different selection to the 
`which.variable` argument. To do this, use the function `MFnest` which takes
the following input arguments:

- `Y`: **MFh** field output as from `MFClusHier()` 
- `which.factor`: As above.

For example:

```{r, message=FALSE}
MFnest(thisMFh, which.factor = "pen")
```

This is the same output as if the user had initially selected for variable "pen":

```{r, message=FALSE}
MFClusHier(formula = lung ~ tx + room/pen/litter, data = a, 
                      which.factor = "pen")
```

# BOOTSTRAPPED MF {-}

## General use. {-}

The function `MFClusBootHier` allows for a bootstrapping approach to calculating
mitigated fraction values. Input arguments are:

*Same as in non-bootstrapped usage*

- `formula`: The formula in form `y ~ x + a/b/c` where `y` is a continuous response, `x` is a factor with two levels of treatment, and `a/b/c` are variables corresponding to the clusters. It is expected that levels of "c" are nested within levels of "b". Nesting is assumed to be in order, left to right, highest to lowest.
- `data`: The data table (`data.frame` or `tibble`) with variables as identified in formula. If there are extra variables, they will be ignored.
- `compare`: A text vector of length two. The first position [1] is treated as the control or reference group two which members of the second position [2] are compared. If no value for the compare vector is supplied, it is assumed that the comparisons are between the `con` and `vac` treatment groups.
- `which.factor`: One or more variable(s) of interest. This can be any of the variables from the data set. If none or NULL is specified, MF will be calculated for the whole tree.

*Additional arguments*

- `nboot`: Number of bootstrapping events.
- `boot.unit`: Boolean whether to sample observations from within those of the
same core.
- `boot.cluster`: Boolean whether to sample which cores are present. If TRUE,
some trees have all the cores represented in the original data while others only 
have a subset.
- `alpha`: Used to calculate empirical hpd by shortest length interval.
- `seed`: Used to initialize random number generator for reproducibility.

A "core" is the unique combination of variable factors from the data, including
the `compare` designation. In the example used in this document one core would 
be Room W/Pen A/Litter 11/vac and another would be Room Z/Pen F/Litter 22/con.
Further discussion about the bootstrapping algorithm can be found in 
[MF algorithms for hierarchical data]. 

## Common coding examples. {-}

Default case, looking only at whole tree and bootstrapping both at the cluster
and unit levels:

```{r, results='hide'}
MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                             data = a)
```

Specifying what bootstrapping sampling to occur:
```{r, results = 'hide'}
MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                              data = a, boot.unit = TRUE, 
                              boot.cluster = FALSE, which.factor = 'room')
```

Adjusting the number of bootstrapping events and alpha:
```{r, results = 'hide', message=FALSE}
MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                                  data = a, boot.unit = FALSE, 
                                  boot.cluster = TRUE,  alpha = 0.1,
                                  which.factor = c('room', 'litter', 'All'))
```

## Basic output. {-}

The default display of `MFClusBootHier` is a table with one row for each unique
mitigated fraction calculated, just like in the non-bootstrapped approach. 
However, instead of summary statistics about the calculated mitigated fraction,
there are values summarizing the mitigated fraction from a bootstrapped
population.

```{r, message = FALSE}
mfboot_multiple <- MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                                  data = a, boot.unit = FALSE, 
                                  boot.cluster = TRUE,  alpha = 0.1,
                                  which.factor = c('room', 'litter', 'All'))
mfboot_multiple
```

\clearpage

The variable columns are:

- `variable`: Which variable was considered when evaluating MF for the row.
- `level`: A unique factor level of the variable which was considered when 
evaluating MF for the row.
- `median`: Median of mitigated fractions calculated from the bootstrapped 
population.
- `etlower`: Lower value of equal tailed range of mitigated fractions calculated 
from the bootstrapped population.
- `etupper`: Upper value of equal tailed range of mitigated fractions calculated 
from the bootstrapped population.
- `htlower`: Lower value of the high tailed range of mitigated fractions calculated 
from the bootstrapped population.
- `htupper`: Upper value of the high tailed range of mitigated fractions calculated 
from the bootstrapped population.
- `mf.obs`: Mitigated fraction value calculated from data input, without 
bootstrapping.

## Advanced usage. {-}

This section covers the technical description of how to access **MFhBoot** and 
`MFnestBoot()` function. For a full discussion of the algorithms of how **MFhBoot**, 
`MFnestBoot()` and `MFClusBootHier` are related, or details of the bootstrapping
algorithm, see [MF algorithms for hierarchical data].

### Bootstrapping step output. {-}

The bootstrapping stage is the most computationally intensive, so a user may
wish to bypass this step subsequently if the only change is what values are 
being passed to the `which.factor` argument. Access bootstrapping step outputs
using the **MFhBoot** field of the output object. For example:

```{r, message=FALSE}
thisBootMFh <- mfboot_multiple$MFhBoot
thisBootMFh
```

This output is a list of four objects:

- `bootmfh`: The rank table for each of the bootstrapped data sets. Formatted
as the rank table for non-bootstrapped data with designation of bootstrapping 
instance.
- `clusters`: Unique cores in the data.
- `compare`: User-supplied value passed to compare argument.
- `mfh`: The rank table from user-supplied data. Formatted as for non-bootstrapped data.

### Reproducibility. {-}

To bootstrap from the same data in a reproducible manner, numerical 
value for the **seed** argument in `MFClusBootHier`. If no value for `seed` is 
specified, the function defaults to a value of `sample(1:1e+05, 1)` and 
subsequent iterations may yield different summaries.

```{r, results='hide'}
MFClusBootHier(formula = lung ~ tx + room/pen/litter, 
                              data = a, seed = 150)
```

Since the bootstrapping step is the most computationally intensive, a user may
wish to use the bootstrapped rank table to recalculate mitigated fraction for 
factors of different variables or a different `alpha` for calculating confidence 
intervals. This is possible using the `MFnestBoot` function,
which takes the following input arguments:

- `x`: **MFhBoot** field output as from `MFClusBootHier`.
- `which.factor`: As above.
- `alpha`: As above.

For example:

```{r}
MFnestBoot(thisBootMFh, which.factor = c('pen', 'All'), alpha = 0.1)
```

# APPENDIX {-}

## Code for example data. {-}

```{r, eval=FALSE}
a <- data.frame(
 room = paste('Room', rep(c('W', 'Z'), each = 24)),
 pen = paste('Pen', rep(LETTERS[1:6], each = 8)),
 litter = paste('Litter', rep(11:22, each = 4)),
 tx = rep(rep(c('vac', 'con'), each = 2), 12),
 stringsAsFactors = FALSE
 )
set.seed(76153)
a$lung[a$tx == 'vac'] <- rnorm(24, 5, 1.3)
a$lung[a$tx == 'con'] <- rnorm(24, 7, 1.3)
a <- a[-48,]
```

## Session details for this manual. {-}

```{r}
sessionInfo()
```




